### 2012-09-19 doc/prelim/document/chapters/introduction/introduction.tex
updating references to zotero database

For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_1993}. The partial differential equation (PDE) community
has focused on various deterministic methods for solutions to linear
problems \citep{saad_2003, kelley_1995}. In between these two areas
are a not widely known small group of stochastic methods for solving
sparse linear systems \citep{hammersley_1964, halton_1962,
  halton_1994}. In recent years, these methods have been further
developed for transport problems in the form of Monte Carlo
Synthetic-Acceleration (MCSA) \citep{evans_2003, evans_2009} but have
yet to be applied to more general sparse linear systems commonly
generated by the computational physics community. Compared to other
methods in these regime, MCSA offers three attractive qualities; (1)
the linear problem operator need not be symmetric or
positive-definite, thereby reducing preconditioning complexity, (2)
the stochastic nature of the solution method provides a natural
solution to the issue of resiliency, and (3) is amenable to
parallelization using modern methods developed by the transport
community \citep{wagner_2011}. The developement of MCSA as a general
linear solver and the development of a parallel MCSA method will be
new and unique features of this work. Resiliency will not be
addressed.

In addition to linear solver advancements, nonlinear solvers may also
benefit from a general and parallel MCSA scheme. In the engineering
community, nonlinear problems are often addressed by either
linearizing the problem or building a segregated scheme and using
traditionaly iterative or direct methods to solve the resulting
system \citep{tannehill_1997}. In the mathematics community, various
Newton methods have been popular \citep{kelley_1995}. Recently,
Jacobian-Free Newton-Krylov (JFNK) schemes \citep{knoll_2004} have
been utilized in multiple physics architectures and advanced single
physics codes \citep{gaston_2009}. The benefits of JFNK schemes are
that the Jacobian is never formed, simplifying the implementation, and
a Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
providing excellent convergence properties for well-conditioned and
well-scaled systems. However, there are two potential drawbacks to
these methods for high fidelity predictive simulations: (1) the
Jacobian is approximated by a first-order differencing method on the
order of machine precision such that this error can grow beyond that
of those in a fine-grained system and (2) for systems that are not
symmetric positive-definite (which will be the case for most
multiphysics systems and certainly for most preconditioned systems)
the Krylov subspace generated by the GMRES solver may become
prohibitively large. To address these issues, this thesis proposes a
new and novel method for nonlinear systems based on the MCSA method.

The Forward-Automated Newton-MCSA (FANM) method is proposed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differencing (FAD) methods
\citep{bartlett_2006}, and MCSA as the inner linear solver. This method
has several attractive properties. First, the first-order
approximation to the Jacobian used in JFNK type methods is eliminated
by generating the Jacobian explicitly with the model equations through
FAD. Second, the Jacobian need not be explicitly formed by the user
but is instead automated through FAD; this eleminates the complexity
of hand-coding derivatives and has also been demonstrated to be more
efficient computationally than evaluating differenced derivatives
\citep{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
subspace during iterations. Although the Jacobian must be explicitly
formed to use MCSA, for problems that take more than a few GMRES
iterations to converge the size of the Krylov subspace will grow
beyond that of the Jacobian. Finally, using MCSA for the linear solve
provides its benefits for preconditioning, potential resiliency, and
parallelism.

>>>>

For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1987}. In between these two areas are a not widely
known small group of stochastic methods for solving sparse linear
systems \citep{hammersley_monte_1964, halton_sequential_1962,
  halton_sequential__1994}. In recent years, these methods have been
further developed for transport problems in the form of Monte Carlo
Synthetic-Acceleration (MCSA)
\citep{t.m._evans__t.j._urbatsch_h._lichtenstein_j.e._morel_residual_2003,
  evans_monte_2009} but have yet to be applied to more general sparse
linear systems commonly generated by the computational physics
community. Compared to other methods in these regime, MCSA offers
three attractive qualities; (1) the linear problem operator need not
be symmetric or positive-definite, thereby reducing preconditioning
complexity, (2) the stochastic nature of the solution method provides
a natural solution to the issue of resiliency, and (3) is amenable to
parallelization using modern methods developed by the transport
community
\citep{j.c._wagner**_s.w._mosher_t.m._evans_d.e._peplow_and_j.a._turner_hybrid_2010}. The
developement of MCSA as a general linear solver and the development of
a parallel MCSA method will be new and unique features of this
work. Resiliency will not be addressed.

In addition to linear solver advancements, nonlinear solvers may also
benefit from a general and parallel MCSA scheme. In the engineering
community, nonlinear problems are often addressed by either
linearizing the problem or building a segregated scheme and using
traditionaly iterative or direct methods to solve the resulting system
\citep{pletcher_computational_1997}. In the mathematics community,
various Newton methods have been popular
\citep{kelley_iterative_1987}. Recently, Jacobian-Free Newton-Krylov
(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
multiple physics architectures and advanced single physics codes
\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that the
Jacobian is never formed, simplifying the implementation, and a Krylov
solver is leveraged (typically GMRES or Conjugate Gradient), providing
excellent convergence properties for well-conditioned and well-scaled
systems. However, there are two potential drawbacks to these methods
for high fidelity predictive simulations: (1) the Jacobian is
approximated by a first-order differencing method on the order of
machine precision such that this error can grow beyond that of those
in a fine-grained system and (2) for systems that are not symmetric
positive-definite (which will be the case for most multiphysics
systems and certainly for most preconditioned systems) the Krylov
subspace generated by the GMRES solver may become prohibitively
large. To address these issues, this thesis proposes a new and novel
method for nonlinear systems based on the MCSA method.

The Forward-Automated Newton-MCSA (FANM) method is proposed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differencing (FAD) methods
\citep{hutchison_automatic_2006}, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated by generating the Jacobian explicitly with the model
equations through FAD. Second, the Jacobian need not be explicitly
formed by the user but is instead automated through FAD; this
eleminates the complexity of hand-coding derivatives and has also been
demonstrated to be more efficient computationally than evaluating
differenced derivatives \citep{hutchison_automatic_2006}. Third,
unlike GMRES, MCSA does not build a subspace during
iterations. Although the Jacobian must be explicitly formed to use
MCSA, for problems that take more than a few GMRES iterations to
converge the size of the Krylov subspace will grow beyond that of the
Jacobian. Finally, using MCSA for the linear solve provides its
benefits for preconditioning, potential resiliency, and parallelism.
